---
title: "CS02 - Predicting Annual Air Pollution"
author: "Michael Tang, Vivian Lee, Jennifer Kim, Madeleine Jimenez, and Sean Pao"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

## Introduction

```{r setup, include=FALSE}
# control global Rmd chunk settings
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Questions
- With what accuracy can we predict US annual average air pollution concentrations?
- Extension question

### Load packages

```{r load-packages, message=FALSE}
library(tidyverse)
library(tidymodels)
library(olsrr)
```

## The Data

Our data comes from the US Environmental Protection Agency (EPA), the National Aeronautics and Space Administration (NASA), the US Census, and the National Center for Health Statistics (NCHS).

It contains 48 features (variables) with values for each of the 876 monitors (observations).

&nbsp;

##### The features are:

 - **id** | Monitor number
   - The county number is indicated before the decimal and the monitor number is indicated after the decimal  
   - Example: 1073.0023 is Jefferson county (1073) and .0023 one of 8 monitors
 - **fips** | Federal information processing standard number for the county where the monitor is located
   - 5 digit id code for counties (zero is often the first value and sometimes is not shown)
   - First two numbers indicate the state, last three numbers indicate the county 
   - Example: Alabama’s state code is 01 because it is first alphabetically
   - Note: Alaska and Hawaii are not included because they are not part of the contiguous US
 - **Lat** | Latitude of the monitor in degrees
 - **Lon** | Longitude of the monitor in degrees
 - **state** | State where the monitor is located
 - **county** | County where the monitor is located
 - **city** | City where the monitor is located
 - **CMAQ** | Estimated values of air pollution from a computational model called [Community Multiscale Air Quality (CMAQ)](https://www.epa.gov/cmaq)
   - A monitoring system that simulates the physics of the atmosphere using chemistry and weather data to predict the air pollution
   - Data from EPA
 - **zcta** | [Zip Code Tabulation Area](https://en.wikipedia.org/wiki/ZIP_Code_Tabulation_Area) where the monitor is located
   - Postal Zip codes are converted into “generalized areal representations” that are non-overlapping
   - Data from the 2010 Census
 - **zcta_area** | Land area of the zip code area in meters squared
   - Data from the 2010 Census
 - **zcta_pop** | Population in the zip code area
   - Data from the 2010 Census
 - **imp_a500, imp_a1000, imp_a5000, imp_a10000, imp_a15000** | Impervious surface measure
   - Impervious surface are roads, concrete, parking lots, buildings
   - Measured within a circle with a radius of 500, 1000, 5000, 10000, and 15000 meters respectively around the monitor
 - **county_area** | Land area of the county of the monitor in meters squared
 - **county_pop** | Population of the county of the monitor
 - **Log_dist_to_prisec** | Log (Natural log) distance to a primary or secondary road from the monitor
   - Highway or major road
 - **log_pri_length_5000, log_pri_length_10000, log_pri_length_15000, log_pri_length_25000** | Count of primary road length in meters in a circle with a radius of 5000, 10000, 15000, and 25000 meters respectively around the monitor (Natural log)
   - Highways only
 - **log_prisec_length_500, log_prisec_length_1000, log_prisec_length_5000, log_prisec_length_10000, log_prisec_length_15000, log_prisec_length_25000** | Count of primary and secondary road length in meters in a circle with a radius of 500, 1000, 5000, 10000, 15000, 25000 meters around the monitor (Natural log)
   - Highway and secondary roads
 - **log_nei_2008_pm25_sum_10000, log_nei_2008_pm25_sum_15000, log_nei_2008_pm25_sum_25000** | Tons of emissions from major sources data base (annual data) sum of all sources within a circle with a radius of 10000, 15000, 25000 meters of distance respectively around the monitor (Natural log)
   - Fine Particulate Matter (diameter 2.5 µm) 
 - **log_nei_2008_pm10_sum_10000, log_nei_2008_pm10_sum_15000, log_nei_2008_pm10_sum_25000** | Tons of emissions from major sources data base (annual data) sum of all sources within a circle with a radius of 10000, 15000, 25000 meters of distance respectively around the monitor (Natural log)
   - Large Coarse Particulate Matter (diameter 10 µm) 
 - **popdens_county** | Population density (number of people per kilometer squared area of the county)
 - **popdens_zcta** | Population density (number of people per kilometer squared area of zcta)
 - From the Census:
   - **nohs** | Percentage of people in zcta area where the monitor is that do not have a high school degree
   - **somehs** | Percentage of people in zcta area where the monitor whose highest formal educational attainment was some high school education
   - **hs** | Percentage of people in zcta area where the monitor whose highest formal educational attainment was completing a high school degree
   - **somecollege** | Percentage of people in zcta area where the monitor whose highest formal educational attainment was completing some college education
   - **associate** | Percentage of people in zcta area where the monitor whose highest formal educational attainment was completing an associate degree
   - **bachelor** |  Percentage of people in zcta area where the monitor whose highest formal educational attainment was a bachelor’s degree
   - **grad** | Percentage of people in zcta area where the monitor whose highest formal educational attainment was a graduate degree
   - **pov** | Percentage of people in zcta area where the monitor is that lived in [poverty](https://aspe.hhs.gov/2008-hhs-poverty-guidelines) in 2008
 - **hs_orless** | Percentage of people in zcta area where the monitor whose highest formal educational attainment was a high school degree or less (sum of nohs, somehs, and hs)
 - **urc2013, urc2006** | [2013](https://www.cdc.gov/nchs/data/series/sr_02/sr02_166.pdf), [2006 Urban-rural classification](https://www.cdc.gov/nchs/data/series/sr_02/sr02_154.pdf) of the county where the monitor is located
   - 6 category variable - 1 is totally urban 6 is completely rural
   - Data from the [National Center for Health Statistics](https://www.cdc.gov/nchs/index.htm)
 - **aod** | Aerosol Optical Depth measurement from a NASA satellite
   - Based on the diffraction of a laser
   - Used as a proxy of particulate pollution
   - Unit-less - higher value indicates more pollution
   - Data from NASA
   
   
### Data Import

```{r}
pm <- read_csv("data/pm25_data.csv")
```

### Data Wrangling

```{r}

```

## Analysis

```{r}

```


## Results & Discussion 

## Conclusion
